---
title: "Estimate *g*(0) and *ESW* with Auxillary Data"
editor: 
  markdown: 
    wrap: 72
---

# A. Process Previous Survey DAS Data

```{r}
#| eval: false

library(tidyverse)
library(Distance)
library(lubridate)
library(here)
library(swfscDAS)
library(MASS)
library(mvnfast)
library(sf)
library(picMaps)

localwd <- here("_R_code","task_2_est_g0_esw")
sp.code <- "033"
```

## Read in previous survey data

```{r}
#| eval: false

#read in input files
das_file <- file.path(localwd, "AllSurveys_g0_ESW_1986-2023.das")
y.proc <- swfscDAS::das_process(das_file)
strata_file_dir <- file.path(localwd, "Strata Files", "region_filter_esw_g0.csv")
```

## Process DAS file and filter for false killer whale regions

```{r}
#| eval: false

y.eff <- swfscDAS::das_effort(
  y.proc, method = "equallength",
  seg.km = 10,
  dist.method = "greatcircle",
  num.cores = 1,
  strata_files = strata_file_dir)
y.eff.sight <- swfscDAS::das_effort_sight(y.eff, sp.codes = sp.code, sp.events = "S")
y.eff.sight.strata <- swfscDAS::das_intersects_strata(y.eff.sight, list(InPoly = strata_file_dir)) 
```

```{r}
#| eval: false
#| echo: false

# Plot code for later...

seg<-y.eff.sight.strata$segdata%>%
  st_as_sf(coords = c("mlon","mlat"))%>%
  st_set_crs(4326)%>%
  st_shift_longitude()
region<-readRDS(file.path(localwd, "Strata Files", "region_filter_poly_esw_g0.rds"))%>%
  st_make_valid()%>%
  st_shift_longitude()

ggplot()+
  theme_classic()+
  geom_sf(data = seg, color = "blue")+
  geom_sf(data = region, color = "black", fill = NA)+
  geom_sf(data = seg[seg$InPoly==1,], color = "red")+
  geom_sf(data = picMaps::hawaii_coast(), fill = "grey")+
  geom_sf(data = picMaps::hawaii_eez(), fill = NA,color = "grey")

```

## Export data for use in *g*(0) and ESW estimation

```{r}
#| eval: false

input_dat<-list()
input_dat$esw.dat <- y.eff.sight.strata$sightinfo %>% 
  filter(SpCode == sp.code, InPoly == 1, OnEffort == TRUE)
input_dat$g0.dat <- y.eff.sight.strata$segdata %>% filter(InPoly == 1, EffType == "S")

saveRDS(input_dat, file.path(localwd, "output","esw_g0_input.rds"))
```

# B. Estimate ESW and *g*(0) Relative to Beaufort Sea State

If you are not continuing from step A in the same session, the following
code must be executed to load the appropriate packages and processed
data from step A. Otherwise, this code chunk can be skipped.

```{r}
#| eval: true
#| output: false

#' load packages
library(tidyverse)
library(Distance)
library(lubridate)
library(here)
library(mvnfast)
library(picMaps)
library(future)
library(doFuture)

localwd <- here("_R_code","task_2_est_g0_esw")
sp.code <- "033"


#' Recall saved data from step 1
input_dat <- readRDS(file.path(localwd, "output","esw_g0_input.rds"))
```

## Estimate ESW relative to Beaufort sea state

Here we are using the `distance` package to fit a half-normal detection
function: $$
p_i(x|\boldsymbol{\theta}_d) = \exp(-x/2\sigma_i^2)
$$ where the detection range parameter is a function of the Beaufort sea
state, $b_i$, on survey segment $i$, $$
\log \sigma_i = \theta_{d,0} + \theta_{d,b} b_i.
$$

```{r}
#| eval: true
#| output: true

df <- data.frame(distance = input_dat$esw.dat$PerpDistKm, beaufort = input_dat$esw.dat$Bft)
df <- df[complete.cases(df),]

detfun <- ds(df, formula = ~beaufort, truncation = 5.5, key="hn") 
summary(detfun)
plot(detfun, showpoints = TRUE)
```

### Estimate ESW for all segments

In this step, the effective strip width ($ESW_i$) is calculated for each
survey segment, $$
ESW_i = \int_0^wp_i(x|\boldsymbol{\theta}_d)dx
$$ where $w$ is the truncation width of the transects, here it is 5.5 km
as can be seen in the previous code. The $ESW$ is the width of the strip
that would result from a survey with perfect detection. Now the fitted
$ESW$ are saved for future if the user is not continuing on to estimate
$g(0;b_i)$, the probability of detection on the trackline when Beaufort
sea state equals $b_i$.

```{r}
#| output: false

modeling_data <- readRDS(file.path(here(), "_R_code", "task_1_segment_das_data", "output","seg_sight_out.rds"))
modeling_data$segdata <- modeling_data$segdata %>% rename(beaufort = avgBft)
modeling_data$segdata[paste0("esw","_",sp.code)] <- predict(detfun, modeling_data$segdata, esw=TRUE)[["fitted"]]
```

## Estimate *g*(0) relative to Beaufort sea state

Now we can use the $ESW$ results to fit a very general DSM model over
false killer whale detections. The density model includes a Beaufort sea
state effect so we can assess the relative density of detections in
different sea states. By assuming that detection along the trackline is
$g(0;b)=1$ for $b = 0$ or $1$ (the calmest seas), we can estimate
trackline detection probability during false killer whale surveys. For
the DSM model, we need to add the effective survey area, $$
ESA_i = 2 \times ESW_i \times L_i
$$that is used to assess the observed density, $y_i$ of false killer
whales for each segment. Then a GAM is fitted to the observed densities
with log expected value: $$
\log E(y_i) = \log ESA_i + f(b_i) + f(Lon_i,Lat_i) + f(year_i) 
$$ where the $f(\dots)$ terms are nonparametric function effects. To
enforce the conditions that $f(b)$ should be monotonically decreasing as
sea state increases, the `scam` package is used instead of `mgcv`. When
the smooth formulation `bs='mpd'` is used, this monotonicity is
enforced.

```{r}
#| output: false

g0_dat <- input_dat$g0.dat
g0_dat$resp <- as.logical(g0_dat[,paste0("nSI_",sp.code)])
g0_dat <- g0_dat%>%dplyr::select(resp, mlat, mlon, avgBft, year,dist) %>% 
  rename(beaufort = avgBft)
g0_dat <- g0_dat[complete.cases(g0_dat), ]

g0_dat$esa <- predict(detfun, newdata = g0_dat, esw=TRUE)[["fitted"]] * 2 * g0_dat$dist
g0_dat <- filter(g0_dat, esa>0)

#' Combine beaufort=0,1 as the same g(0) = 1
g0_est<-scam::scam(formula = resp ~ s(I(pmax(beaufort-1,0)), bs='mpd', k=4)
                     + s(mlat, mlon, bs='tp') + s(year, k=4), 
                   offset = log(esa+0.0001),
                   family=poisson,
                   data=g0_dat,
                   not.exp=TRUE,
                   gamma=1.4)
```

### Estimate relative *g*(0) for each Beaufort sea state

Now that the model is fit, we can estimate $g(0;b)$ by examining the
relative expected density for any specific segment under different
Beaufort conditions. Because the $f(b_i)$ term is additive, it does not
matter which is used because they will all be the same. So, we just
formulate a generic segment with varying Beaufort states. For the
predicted density, $$
\mu_b = E(y_*) = \exp( \log ESA_* + f(b) + f(Lon_*,Lat_*) + f(year_*)),
$$ the estimated trackline probability is $$
g(0;b) = \mu_b / \mu_0
$$

```{r}
newdata_g0 <- data.frame(beaufort = 0:7,
                         mlat = mean(modeling_data$segdata$mlat, na.rm=TRUE),
                         mlon = mean(modeling_data$segdata$mlon, na.rm=TRUE),
                         year = round(mean(modeling_data$segdata$year, na.rm=TRUE)))

ps <- scam::predict.scam(g0_est, newdata = newdata_g0, type="response")
Rg0 <- ps/ps[1]
names(Rg0) <- 0:7
round(Rg0,2) # for display only
```

### Add *g*(0) estimate to all segments and save final modeling data set

To account for survey segments with non-integer Beaufort states, because
it changed over the course of the segment, the $g(0;b)$ value was
linearly interpolated between the adjacent $b$ values. Then model data
for the current survey is exported for use in the pelagic false killer
whale DSM. The $g(0;b_i)$ and $ESW_i$ will serve as an effort offset.

```{r}
#| output: false

BFlow <- floor(modeling_data$segdata$beaufort + 1)
BFhigh <- ceiling(modeling_data$segdata$beaufort + 1)
w <- BFhigh - (modeling_data$segdata$beaufort + 1)
modeling_data$segdata[,paste0("g0_w_",sp.code)] <- w
modeling_data$segdata[,paste0("g0","_",sp.code)] <- w*Rg0[BFlow] + (1-w)*Rg0[BFhigh]

saveRDS(modeling_data, file.path(localwd, "output","seg_sight_out_g0_ESW.rds"))
```

# C. Monte Carlo Estimation of ESW and *g*(0) Parameter Covariance Matrix

To propagate the uncertainty for effort terms into the false killer
whale DSM model, the covariance matrix for the detection terms needs to
be estimated because the $g(0;b)$ model depends on the $ESA_i$ values
calculated from the distance model. Therefore, the most straightforward
method to account for the covariance between the distance sampling
parameters, $\theta_d$ and the $g(0,b)$ parameters, $\theta_g$, is to
perform a Monte Carlo simulation. First, we sample 5,000 $\theta_d$
replications from its large sample distribution,
$N(\hat{\theta}_d, V_d)$, where $\hat{\theta}_d$ are the estimated
distance function parameters and $V_d$ is the estimated covariance
matrix. For each replication, we calculate $ESA_i$ for the survey
segments and fit the scam model to estimate the $f(b)$ parameters,
$\theta_g$. The full estimated covariance matrix, $V_\theta$, for =
$\hat{\theta}=(\hat{\theta}_d, \hat{\theta}_g)$, was estimated with the
empirical covariance matrix of the 5,000 samples.

**Warning!!!** The following Monte Carlo code takes a very long time to
run for the false killer whale analysis. To improve speed of the
simulation, we have looped over blocks of smaller simulation in
parallel. For the code below we use `npar <- 10` separate `R` instances,
each sampling `samps <- 500` $\theta$ values for a total of 5,000
$\theta$ replications. If you want to see the code in action, we suggest
that you dramatically reduce the overall sample size to assess how long
it will take the full sample to execute.

```{r}
#| eval: false

# Design matrix for g(0) estimates 
Lp <- scam::predict.scam(g0_est, newdata = newdata_g0, type="lpmatrix")
# Subset for columns affecting g(0):
bft_idx <- grep("beaufort", colnames(Lp))
Lp <- Lp[,bft_idx]

# Draw detection parameter sample:
detfun_g0 <- detfun
v_theta <- solve(detfun_g0$ddf$hessian)
pars.esw <- detfun_g0$ddf$par
```

```{r}
#| eval: false
npar <- 10
samps <- 500 # total samples = npar * samps

plan("multisession", workers=npar)
set.seed(8675309)

g0_par_sim <- foreach(i = 1:npar,  .options.future = list(seed = TRUE), .errorhandling = "pass") %dofuture% {
  theta_star <- matrix(NA, samps, length(detfun$ddf$par))
  v_alpha_star <- matrix(NA, samps, length(bft_idx)^2)
  alpha_star <- matrix(NA, samps, length(bft_idx))
  Rg0_star <- matrix(NA, samps, 8)
  # Constructing variance-covariance matrix 
  for(j in 1:samps){
    g0_star_out <- sim_esw_g0(pars.esw, v_theta, detfun, g0_dat)
    g0_star <- g0_star_out$g0_star
    theta_star[j,] <-  g0_star_out$theta_star
    v_alpha_star[j,] <- as.vector(g0_star$Vp.t[bft_idx, bft_idx])
    alpha_star[j,] <- g0_star$coefficients.t[bft_idx]
    Rg0_star[j,] <- g0_star_out$Rg0
    
  }
  list(theta_star=theta_star, alpha_star=alpha_star, v_alpha_star=v_alpha_star, Rg0_star=Rg0_star)
}
plan("sequential")
```

Here, we are combining the parallel samples back together and
calculating the covatiance matrix $V_\theta$. Then, we export the
results for use in the abundance prediction uncertainty assessment
stage.

```{r}
#| eval: false

theta_star <- lapply(g0_par_sim, \(x) x$theta_star) %>% do.call(rbind,.)
alpha_star <- lapply(g0_par_sim, \(x) x$alpha_star) %>% do.call(rbind,.)
v_alpha_star <- lapply(g0_par_sim, \(x) x$v_alpha_star) %>% do.call(rbind,.)
Rg0_star <- lapply(g0_par_sim, \(x) x$Rg0_star) %>% do.call(rbind,.)

## Compose full variance/covariance from law of total variance
v_alpha <- colMeans(v_alpha_star) %>% matrix(.,sqrt(length(.))) +
  var(alpha_star)
c_alpha_theta <- cov(alpha_star, theta_star)
V <- rbind(
  cbind(v_alpha, c_alpha_theta),
  cbind(t(c_alpha_theta), var(theta_star))
)

varprop_list <- list(
  newdata_g0 = newdata_g0,
  Lp_g0 = Lp,
  V_par = V,
  par_sim = cbind(alpha_star, theta_star),
  Rg0_sim = Rg0_star,
  detfun = detfun, 
  g0_dat = g0_dat,
  g0_est = g0_est,
  bft_idx = bft_idx
)

saveRDS(varprop_list, file = file.path(datadir, "output","varprop_list.rds"))
```
